Finally Tagless, Partially Evaluated

### 2008-09-10 tagless/tagless-final.tex
Fixed formalization in Section 3

Comparing |Symantics| with Fig.~\ref{fig:object}
shows how to represent \emph{every} well-typed object term in the
metalanguage.
We formalize this representation by defining $H$ and~$M$,
two inductive maps from terms and types in our object language
to terms and types in Haskell and OCaml:
\begin{align}
    H(\ZZ) &= \texttt{Int} &
    M(\ZZ) &= \texttt{int} \notag\\
    H(\BB) &= \texttt{Bool} &
    M(\BB) &= \texttt{bool} \notag\\
    H(t_1 \to t_2) &= H(t_1) \mathbin{\texttt{->}} H(t_2) &
    M(t_1 \to t_2) &= M(t_1) \mathbin{\texttt{->}} M(t_2) \displaybreak[0] \\
    H(x) &= x &
    M(x) &= x \notag\\
    H(\fun{x}e) &= \texttt{lam (\textbackslash x -> $H(e)$)} &
    M(\fun{x}e) &= \texttt{lam (fun x -> $M(e)$)} \notag\\
    H(\fix{f}e) &= \texttt{fix (\textbackslash x -> $H(e)$)} &
    M(\fix{x}e) &= \texttt{fix (fun x -> $M(e)$)} \notag\\
    H(e_1 e_2) &= \texttt{app $H(e_1)$ $H(e_2)$} &
    M(e_1 e_2) &= \texttt{app $M(e_1)$ $M(e_2)$} \notag\\
    H(n) &= \texttt{int $n$} &
    M(n) &= \texttt{int $n$} \notag\\
    H(\True) &= \texttt{bool True} &
    M(\True) &= \texttt{bool true} \notag\\
    H(\False) &= \texttt{bool False} &
    M(\False) &= \texttt{bool false} \notag\\
    \!H(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $H(e)$ $H(e_1)$ $H(e_2)$}} \notag\\
    \!M(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $M(e)$ (fun () -> $M(e_1)$) (fun () -> $M(e_2)$)}} \notag\\
    H(e_1 + e_2) &= \texttt{add $H(e_1)$ $H(e_2)$} &
    M(e_1 + e_2) &= \texttt{add $M(e_1)$ $M(e_2)$} \notag\\
    H(e_1 \times e_2) &= \texttt{mul $H(e_1)$ $H(e_2)$} &
    M(e_1 \times e_2) &= \texttt{mul $M(e_1)$ $M(e_2)$} \notag\\
    H(e_1 \le e_2) &= \texttt{leq $H(e_1)$ $H(e_2)$} &
    M(e_1 \le e_2) &= \texttt{leq $M(e_1)$ $M(e_2)$}
\end{align}
These definitions assume that our object language, Haskell, and OCaml
use the same variable names~$x$ and integer literals~$n$.
If $\Gamma$ is a typing context $x_1:t_1,\dotsc,x_n:t_n$ in the object
language, then we define the metalanguage contexts
\begin{align}
    \texttt{repr\;$H(\Gamma)$}
    &= x_1:\texttt{repr}\;H(t_1),\dotsc,x_n:\texttt{repr}\;H(t_n) \text, \notag\\
    \texttt{('c,\:$M(\Gamma)$)\;repr}
    &= x_1:\texttt{('c,\:$M(t_1)$)\;repr},\dotsc,x_n:\texttt{('c,\:$M(t_n)$)\;repr} \text.
\end{align}

@@

\medskip
\caption{The tagless staged interpreter \texttt{C}}
\label{fig:interpreter-C}
\end{figure}
\fi
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1| with
\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
with
\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
and does not
call the interpreter, unlike the recursive calls to |eval0| and
|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
The resulting code obviously contains no tags and no pattern matching.
The environment classifiers here, like the tuple types in \S\ref{ourapproach},
make it a type error to run an open expression.
\ifshort
The accompanying code shows the Haskell implementation. 
\else
\begin{proposition}
    For all object terms $e$ of object type $t$ (encoded as $e'$),
    evaluating $e'$ via the compiler |C| always terminates and
    generates tagless code (i.e. of type $t$ |code|) free of pattern matching.
\end{proposition}
This is syntactically clear from the body of |C|, and follows
readily by structural induction and the properties%
\footnote{which are not easy to establish \citep{MetaML}} of |.~| and
|.< >.|.

We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it: we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
(Alternatively, we could use Template Haskell as our staging facility:
\texttt{ByteCode} can be mapped to the abstract syntax of Template
Haskell. The output of our compiler will then be assuredly type-correct
Template Haskell.)
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. (The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.) The
accompanying code gives the full details.
\fi

>>>>

Comparing |Symantics| with Fig.~\ref{fig:object}
shows how to represent \emph{every} well-typed object term in the
metalanguage.
We formalize this representation by defining $H$ and~$M$,
two inductive maps from terms and types in our object language
to terms and types in Haskell and OCaml:
\begin{align}
    H(\ZZ) &= \texttt{Int} &
    M(\ZZ) &= \texttt{int} \notag\\
    H(\BB) &= \texttt{Bool} &
    M(\BB) &= \texttt{bool} \notag\\
    H(t_1 \to t_2) &= H(t_1) \mathbin{\texttt{->}} H(t_2) &
    M(t_1 \to t_2) &= M(t_1) \mathbin{\texttt{->}} M(t_2) \displaybreak[0] \\
    H(x) &= x &
    M(x) &= x \notag\\
    H(\fun{x}e) &= \texttt{lam (\textbackslash $x$ -> $H(e)$)} &
    M(\fun{x}e) &= \texttt{lam (fun $x$ -> $M(e)$)} \notag\\
    H(\fix{f}e) &= \texttt{fix (\textbackslash $x$ -> $H(e)$)} &
    M(\fix{x}e) &= \texttt{fix (fun $x$ -> $M(e)$)} \notag\\
    H(e_1 e_2) &= \texttt{app $H(e_1)$ $H(e_2)$} &
    M(e_1 e_2) &= \texttt{app $M(e_1)$ $M(e_2)$} \notag\\
    H(n) &= \texttt{int $n$} &
    M(n) &= \texttt{int $n$} \notag\\
    H(\True) &= \texttt{bool True} &
    M(\True) &= \texttt{bool true} \notag\\
    H(\False) &= \texttt{bool False} &
    M(\False) &= \texttt{bool false} \notag\\
    \!H(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $H(e)$ $H(e_1)$ $H(e_2)$}} \notag\\
    \!M(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $M(e)$ (fun () -> $M(e_1)$) (fun () -> $M(e_2)$)}} \notag\\
    H(e_1 + e_2) &= \texttt{add $H(e_1)$ $H(e_2)$} &
    M(e_1 + e_2) &= \texttt{add $M(e_1)$ $M(e_2)$} \notag\\
    H(e_1 \times e_2) &= \texttt{mul $H(e_1)$ $H(e_2)$} &
    M(e_1 \times e_2) &= \texttt{mul $M(e_1)$ $M(e_2)$} \notag\\
    H(e_1 \le e_2) &= \texttt{leq $H(e_1)$ $H(e_2)$} &
    M(e_1 \le e_2) &= \texttt{leq $M(e_1)$ $M(e_2)$}
\end{align}
These definitions assume that our object language, Haskell, and OCaml
use the same variable names~$x$ and integer literals~$n$.
If $\Gamma$ is a typing context $x_1:t_1,\dotsc,x_n:t_n$ in the object
language, then we define the metalanguage contexts
\begin{align}
    \texttt{repr\;$H(\Gamma)$}
    &= x_1:\texttt{repr}\;H(t_1),\dotsc,x_n:\texttt{repr}\;H(t_n) \text, \\
    \texttt{('c,\:$M(\Gamma)$)\;repr}
    &= x_1:\texttt{('c,\:$M(t_1)$)\;repr},\dotsc,x_n:\texttt{('c,\:$M(t_n)$)\;repr} \text.
\end{align}

@@

\medskip
\caption{The tagless staged interpreter \texttt{C}}
\label{fig:interpreter-C}
\end{figure}
\fi
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1| with
\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
with
\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
and does not
call the interpreter, unlike the recursive calls to |eval0| and
|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
The resulting code obviously contains no tags and no pattern matching.
The environment classifiers here, like the tuple types in \S\ref{ourapproach},
make it a type error to run an open expression.
\ifshort
The accompanying code shows the Haskell implementation. 
\else
\begin{proposition}
If an object term~$e$ has the type~$t$
in the context $x_1:t_1,\dotsc,x_n:t_n$,
then in a MetaOCaml environment
\[
    \texttt{open C},\quad
    x_1 \mapsto \texttt{.<$y_1$>.},\enspace
    \dotsc,\enspace
    x_n \mapsto \texttt{.<$y_n$>.}
\]
where each $y_i$ is a future-stage variable of type $M(t_i)$,
the MetaOCaml term~$M(e)$ evaluates to a code value,
of type~\texttt{('c,\:$M(t)$)\;code} (polymorphic over~|'c|),
that contains no pattern-matching operations.
\end{proposition}
\begin{proof}
    By structural induction on the typing derivation of~$e$.
\end{proof}
\begin{corollary}
If a closed object term~$e$ has type~$t$,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (C)
    \end{tabular}}
\]
the expression \texttt{RESULT.term ()} evaluates to a code value,
of type~\texttt{('c,\:$M(t)$)\;code} (polymorphic over~|'c|),
that contains no pattern-matching operations.
\end{corollary}

We have also implemented this compiler in Haskell. 
Since Haskell
has no convenient facility for typed staging, we emulate
it by defining a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
(Alternatively, we could use Template Haskell \citep{sheard-template} as our staging facility:
\texttt{ByteCode} can be mapped to the abstract syntax of Template
Haskell. The output of our compiler would then be assuredly type-correct
Template Haskell.)
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation is quite similar (but slightly more
verbose) than the MetaOCaml code above. (The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.) The
accompanying code gives the full details.
\fi
